---
title: "BST 222 - Midterm Assignment Codes"
author: "Xiaowei Zeng"
date: "`r Sys.Date()`"
output: pdf_document
# fontsize: 11pt
#output: 
#  pdf_document:
#    number_sections: true
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align='center', out.width='65%')
```

# Here are the codes for problems in the instruction document of Midterm Project, not the codes for the report!

## Preparation
Load the R packages and the dataset.
```{r}
library(KMsurv)
library(survival)

data(burn)
# set variable s to factors
burn0 <- data.frame(
  Obs=burn$Obs,
  T1=burn$T1,
  D1=burn$D1,
  T2=burn$T2,
  D2=burn$D2,
  T3=burn$T3,
  D3=burn$D3,
  Treatment=factor(burn$Z1,labels=c("Routine","Cleansing")),
  Gender=factor(burn$Z2,labels=c("Male","Female")),
  Race=factor(burn$Z3,labels=c("Nonwhite","White")),
  PercentBurn=burn$Z4, 
  SiteHead=factor(burn$Z5,labels=c("NotBurned","Burned")),
  SiteButtock=factor(burn$Z6,labels=c("NotBurned","Burned")),
  SiteTrunk=factor(burn$Z7,labels=c("NotBurned","Burned")),
  SiteUpperLeg=factor(burn$Z8,labels=c("NotBurned","Burned")),
  SiteLowerLeg=factor(burn$Z9,labels=c("NotBurned","Burned")),
  SiteRespTract=factor(burn$Z10,labels=c("NotBurned","Burned")),
  BurnSource=factor(burn$Z11,labels=c("Chemical","Scald","Electric","Flame"))
)
burn <- tmerge(burn0, burn0, Obs, tstop=T3)
# set time to excision
burn <- tmerge(burn, burn0, Obs, excis=tdc(T1))
# set time to antibiotics
burn <- tmerge(burn, burn0, Obs, antib=tdc(T2))
# status at the end, either they had an infection or they were censored
burn$status <- as.integer(with(burn, (tstop == T3 & D3)))
# preview the data
head(burn)
```

Create a survival object with time independent covariates.
```{r}
surv1 <- Surv(burn0$T3, burn0$D3)
```

Create a survival object with time dependent covariates.
```{r}
surv2 <- Surv(time=burn$tstart, time2=burn$tstop,
              event=burn$status, type="counting")
```

Suppose that the significance level in this study is 0.1.

## 1. 

Plot the Kaplan-Meier curves for the treated and untreated patients.
```{r}
KMcurves <- survfit(surv1~Treatment, data=burn0)
plot(KMcurves, col=1:2, xlab="Time (Days)", ylab="Survival", lwd=2)
legend("topright", c("Routine Care","Total Body Cleansing"), col=1:2, lwd=2)
title('KM Survival Curves with Treatment')
```

Use `survdiff` to test for whether the curves are different.
```{r}
survdiff(surv1~Treatment, data=burn0)
```

The $p$-value of the log rank test is 0.05<0.1, indicating that the curves are significantly different.

## 2. 
Plot the cumulative hazards vs. time.
```{r}
NAcurves <- survfit(surv1~Treatment, type="fleming-harrington", data=burn0)
plot(NAcurves, col=1:2, fun="cumhaz", xlab="Time (Days)",
     ylab="Cumulative Hazards", lwd=2)
legend("topleft", c("Routine Care","Total Body Cleansing"), col=1:2, lwd=2)
title('NA Cumulative Hazards Curves with Treatment')
```

The cumulative hazards of Cleansing group are larger than those with routine bathing for nearly all time points (except the first several days).

Plot the complimentary log-log survival vs. log time.
```{r}
plot(NAcurves, col=1:2, fun="cloglog", xlab="Time (Days in log scale)", ylab="Log(Cumulative Hazards)", lwd=2)
legend("topleft", c("Routine Care","Total Body Cleansing"), col=1:2, lwd=2)
title('Complimentary Log-Log Survival Curves')
```

The two lines are roughly parallel, so we can assume that the hazards are proportional.

## 3.

Construct Cox model using only the time-independent predictors. Start with one using only Treatment. 

```{r}
cox1 <- coxph(surv1~Treatment, data=burn0)
summary(cox1)
```

Include all the time independent variables in the initial model. Use stepwise model selection method to decide which varaibles should be included in the model. 

```{r}
cox2 <- coxph(surv1~Treatment+Gender+Race+PercentBurn+SiteHead+SiteButtock+SiteTrunk+
                SiteUpperLeg+SiteLowerLeg+SiteRespTract+BurnSource, data=burn0)
step(cox2)
```

All the burn site variables are excluded from the model. None of them show a significant effect on infection risk. The final model with time independent variables should be fit with Treatment, Gender, Race and BurnSource variable.

```{r}
cox3 <- coxph(surv1~Treatment+Gender+Race+BurnSource, data=burn0)
summary(cox3)
```

## 4. 

Run the usual suite of model checking methods and report any interesting findings. Possibly alter the model as a result.

1. Use `cox.zph` to test PH assumption.
```{r}
burn0.zph <- cox.zph(cox3)
print(burn0.zph)
```

The BurnSource variable has a $p$-value of 0.038<0.1 and the $p$-value for the global test is 0.04<0.1. Thus, it does not satisfy proportionality of hazards assumption.

2. Make the plots vs. time of the Schoenfeld residuals.

```{r }
plot(burn0.zph[1], main = 'Schoenfeld Residuals for Treatment')
plot(burn0.zph[2], main = 'Schoenfeld Residuals for Gender')
plot(burn0.zph[3], main = 'Schoenfeld Residuals for Race')
plot(burn0.zph[4], main = 'Schoenfeld Residuals for BurnSource')
```

Have a look at the patterns of variables vs. time. Severe time dependence!

Plot the KM curve for the event of infection and BurnSource. It shows a significant intersection between the flame curve and the scald curve. We need to employ stratification.
```{r, message=FALSE}
library(survminer)
ggsurvplot(survfit(surv1~BurnSource, data=burn0))
```

First test whether we can use strata (whether thes coefficient of other variables are the same across strata). I fit seperate models, but it cannot converge because the sample size for each source of burn is not balanced.
```{r}
table(burn0$BurnSource)
```

I turn to fit an interaction model and find that all the variables are not significant. We can simply assume that the coefficients are the same across strata.
```{r}
inter.cox <- coxph(surv1~(Treatment+Gender+Race)*strata(BurnSource), data=burn0)
summary(inter.cox)
```

Construct a new stratified model. Without time-dependent variable, this is the final model. 
```{r}
cox4 <- coxph(surv1~Treatment+Gender+Race+strata(BurnSource), data=burn0)
summary(cox4)
```

Plot the stratified model.
```{r}
plot(survfit(cox4, data.frame(Treatment=levels(burn0$Treatment),
                              Gender=levels(burn0$Gender)[1],
                              Race=levels(burn0$Race)[1])), 
     col=1:4, main = 'Stratified Cox Model', ylim=c(0.8, 1))
```

1. Use `cox.zph` to test PH assumption.
```{r}
burn0.zph <- cox.zph(cox4)
print(burn0.zph)
```

The model satisfies PH assumption now because the $p$-value for the global model is 0.192>0.1.

2. Make the plots vs. time of the Schoenfeld residuals.
```{r }
plot(burn0.zph[1], main = 'Schoenfeld Residuals for Treatment')
plot(burn0.zph[2], main = 'Schoenfeld Residuals for Gender')
plot(burn0.zph[3], main = 'Schoenfeld Residuals for Race')
```

Still strong time dependece.

3. Plot the cumulative hazard of the Cox-Snell residuals.
```{r}
burn0.mart <- residuals(cox4, type='martingale')
burn0.cs <- burn0$D3 - burn0.mart
surv.csr <- survfit(Surv(burn0.cs, burn0$D3) ~ 1, type='fleming-harrington')
plot(surv.csr, fun='cumhaz', ylim=c(0, 3)); abline(0, 1)
title('Cumulative Hazard of Cox-Snell Residuals')
```

The line with slope 1 and intercept 0 does not fit the curve well. There may be a lack of fit of goodness in this model.

4. Plot the martingale residuals vs. the linear predictor.
```{r}
burn0.preds <- predict(cox4) # linear predictor
plot(burn0.preds, burn0.mart, xlab='Linear Predictor', ylab='Martingale Residual')
title('Martingale Residuals vs. Linear Predictor')
burn0.mart[burn0.mart < - 1]
```

5. Plot the deviance residuals vs. the linear predictor.
```{r}
burn0.dev <- residuals(cox4, type='deviance') 
plot(burn0.preds, burn0.dev, xlab='Linear Predictor', ylab='Deviance Residual')
title('Deviance Residuals vs. Linear Predictor')
burn0.dev[burn0.dev > 2]
```

6. Plot the three dfbeta values by observation order.
```{r}
burn0.dfb <- residuals(cox4, type='dfbeta')
plot(burn0.dfb[, 1], xlab='Observation Order', ylab='dfbeta for Treatment')
title('dfbeta Values by Observation Order for Treatment')
burn0.dfb[abs(burn0.dfb[, 1]) > 0.05, 1]
```

```{r}
plot(burn0.dfb[, 2], xlab='Observation Order', ylab='dfbeta for Gender')
title('dfbeta Values by Observation Order for Gender')
burn0.dfb[abs(burn0.dfb[, 2]) > 0.1, 2]
```

```{r}
plot(burn0.dfb[, 3], xlab='Observation Order', ylab='dfbeta for Race')
title('dfbeta Values by Observation Order for Race')
burn0.dfb[abs(burn0.dfb[, 3]) > 0.2, 3]
```

The most important observations to examine should be 32, 79, 90, 115, 116, and 153.

```{r}
burn0[c(32, 79, 90, 115, 116, 153), ]
```

- 32: This patient experienced an electric burn both on the head and trunk, indicating a relatively severe burn condition. However, despite not receiving interventions such as body cleansing, surgical excision, or antibiotic treatment, the patient remained infection-free until exiting the study. It is essential to acknowledge that this could be attributed to either his early withdrawal from the study or the early ending of the study.
- 79: This patient belongs to the intervention group and is a female, representing the smaller proportion of females in the overall sample. Additionally, she has a relatively small burned percentage of total surface area. Previous models show that the intervention group has a lower risk of infection compared to the control group, and females, although not statistically significant, tend to have a lower risk as well. However, in her case, she rapidly developed an infection on the second day of entering the study.
- 90: This patient belongs to the intervention group and is a female, representing the smaller proportion of females in the overall sample. She suffered from chemical burn, underwent both surgical excision and antibiotics treatment, but she still got infected relatively early.
- 115: This patient is from the intervention group and rapidly got infected on the third day of entering the study. Considering that the patient has burns in four locations, with a burn percentage of surface area accounting for 50\%, this doesn't appear to be particularly unusual.
- 116: This patient is the only one in the nonwhite group who is infected, and therefore, it will have a significant negative impact on the coefficient of the Race variable.
- 153: Despite being in the intervention group and having a very low burn percentage, this patient rapidly developed an infection on the second day after entering the study.

The majority of these outliers are detected due to unexpectedly early infections. They exhibit characteristics associated with lower infection risk, such as being in the intervention group, having a small burn percentage, and undergoing excision or antibiotic treatment. Despite these factors, they still experienced infections just a few days after entering the study. 

## 5.

Add time-dependent variables into the model and use stepwise method to conduct model selection just the same as before.
```{r}
cox5 <- coxph(surv2~Treatment+Gender+Race+PercentBurn+SiteHead+SiteButtock+SiteTrunk+
                SiteUpperLeg+SiteLowerLeg+SiteRespTract+BurnSource+excis+antib, data=burn)
step(cox5)
```

The best model with the lowest AIC has Treatment, Race, BurnSource and excis (time-dependent) as the covariates.
```{r}
cox6 <- coxph(surv2~Treatment+Race+BurnSource+excis, data=burn)
summary(cox6)
```

## 6.

Run the usual suite of model checking methods and report any interesting findings. Possibly alter the model as a result.

1. Use `cox.zph` to test PH assumption.
```{r}
burn.zph <- cox.zph(cox6)
print(burn.zph)
```

The BurnSource variable has a $p$-value of 0.035<0.1 and the $p$-value for the global test is 0.051<0.1. Thus, it does not satisfy proportionality of hazards assumption.

2. Make the plots vs. time of the Schoenfeld residuals.

```{r }
plot(burn.zph[1], main = 'Schoenfeld Residuals for Treatment')
plot(burn.zph[2], main = 'Schoenfeld Residuals for Race')
plot(burn.zph[3], main = 'Schoenfeld Residuals for BurnSource')
plot(burn.zph[4], main = 'Schoenfeld Residuals for Excision')
```

We need to employ stratification. First test whether we can use strata (whether thes coefficient of other variables are the same across strata). I fit seperate models, but it cannot converge because the sample size for each source of burn is not balanced. I turn to fit an interaction model and find that all the variables are not significant. We can simply assume that the coefficients are the same across strata.
```{r}
inter.cox <- coxph(surv2~(Treatment+Race+excis)*strata(BurnSource), data=burn)
summary(inter.cox)
```

Construct a new stratified model. This is the final model including time-dependent variable.
```{r}
cox7 <- coxph(surv2~Treatment+Race+excis+strata(BurnSource), data=burn)
summary(cox7)
```

Plot the stratified model.
```{r}
plot(survfit(cox7, data.frame(Treatment=levels(burn$Treatment),
                              Race=levels(burn$Race)[1], 
                              excis=1)), 
     col=1:4, main = 'Stratified Cox Model', ylim=c(0.85, 1))
```

1. Use `cox.zph` to test PH assumption.
```{r}
burn.zph <- cox.zph(cox7)
print(burn.zph)
```

The model satisfies PH assumption now because the $p$-value for the global model is 0.288>0.1.

2. Make the plots vs. time of the Schoenfeld residuals.
```{r }
plot(burn.zph[1], main = 'Schoenfeld Residuals for Treatment')
plot(burn.zph[2], main = 'Schoenfeld Residuals for Race')
plot(burn.zph[3], main = 'Schoenfeld Residuals for Excision')
```

Patterns with time still exists.

3. Plot the cumulative hazard of the Cox-Snell residuals.
```{r}
burn.mart <- residuals(cox7, type='martingale')
burn.cs <- burn$status - burn.mart
surv.csr <- survfit(Surv(burn.cs, burn$status) ~ 1, type='fleming-harrington')
plot(surv.csr, fun='cumhaz', ylim=c(0, 2)); abline(0, 1)
title('Cumulative Hazard of Cox-Snell Residuals')
```

The line with slope 1 and intercept 0 fits the curve relatively well, so we don’t see lack of fit based on these residuals. This model is much better than the Model 4.

4. Plot the martingale residuals vs. the linear predictor.
```{r}
burn.preds <- predict(cox7) # linear predictor
plot(burn.preds, burn.mart, xlab='Linear Predictor', ylab='Martingale Residual')
title('Martingale Residuals vs. Linear Predictor')
burn$Obs[burn.mart < - 1]
```

5. Plot the deviance residuals vs. the linear predictor.
```{r}
burn.dev <- residuals(cox7, type='deviance') 
plot(burn.preds, burn.dev, xlab='Linear Predictor', ylab='Deviance Residual')
title('Deviance Residuals vs. Linear Predictor')
burn$Obs[burn.dev > 2]
```

6. Plot the three dfbeta values by observation order.
```{r}
burn.dfb <- residuals(cox7, type='dfbeta')
plot(burn.dfb[, 1], xlab='Observation Order', ylab='dfbeta for Treatment')
title('dfbeta Values by Observation Order for Treatment')
burn$Obs[abs(burn.dfb[, 1]) > 0.05]
```

```{r}
plot(burn.dfb[, 2], xlab='Observation Order', ylab='dfbeta for Race')
title('dfbeta Values by Observation Order for Race')
burn$Obs[abs(burn.dfb[, 2]) > 1]
```

```{r}
plot(burn.dfb[, 3], xlab='Observation Order', ylab='dfbeta for Excision')
title('dfbeta Values by Observation Order for Excision')
burn$Obs[abs(burn.dfb[, 3]) > 0.1]
```

The most important observations to examine should be 32, 79, 90, 115, 116, and 153.

```{r}
burn0[c(32, 79, 115, 116, 125, 153), ]
```

- 32: This patient experienced an electric burn both on the head and trunk, indicating a relatively severe burn condition. However, despite not receiving interventions such as body cleansing, surgical excision, or antibiotic treatment, the patient remained infection-free until exiting the study. It is essential to acknowledge that this could be attributed to either his early withdrawal from the study or the early ending of the study.
- 79: This patient belongs to the intervention group and is a female, representing the smaller proportion of females in the overall sample. Additionally, she has a relatively small burned percentage of total surface area. Previous models show that the intervention group has a lower risk of infection compared to the control group, and females, although not statistically significant, tend to have a lower risk as well. However, in her case, she rapidly developed an infection on the second day of entering the study.
- 115: This patient is from the intervention group and rapidly got infected on the third day of entering the study. Considering that the patient has burns in four locations, with a burn percentage of surface area accounting for 50\%, this doesn't appear to be particularly unusual.
- 116: This patient is the only one in the nonwhite group who is infected, and therefore, it will have a significant negative impact on the coefficient of the Race variable.
- 125: Despite being in the intervention group, having a relatively low burn percentage, and receiving antibiotic treatment before infection, this patient rapidly developed an infection on the third day after entering the study.
- 153: Despite being in the intervention group and having a very low burn percentage, this patient rapidly developed an infection on the second day after entering the study.

The majority of these outliers are detected due to unexpectedly early infections. They exhibit characteristics associated with lower infection risk, such as being in the intervention group, having a small burn percentage, and undergoing excision or antibiotic treatment. Despite these factors, they still experienced infections just a few days after entering the study. 

