---
title: "BST 222 - Homework 5"
author: "Xiaowei Zeng"
date: "`r Sys.Date()`"
output: pdf_document
# fontsize: 11pt
#output: 
#  pdf_document:
#    number_sections: true
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.align='center', out.width='65%')
```

## Preparation
Load the R packages and the dataset.
```{r}
library(KMsurv)
library(survival)

data(burn)
# set variable s to factors
burn0 <- data.frame(
  Obs=burn$Obs,
  T1=burn$T1,
  D1=burn$D1,
  T2=burn$T2,
  D2=burn$D2,
  T3=burn$T3,
  D3=burn$D3,
  Treatment=factor(burn$Z1,labels=c("Routine","Cleansing")),
  Gender=factor(burn$Z2,labels=c("Male","Female")),
  Race=factor(burn$Z3,labels=c("Nonwhite","White")),
  PercentBurn=burn$Z4, 
  SiteHead=factor(burn$Z5,labels=c("NotBurned","Burned")),
  SiteButtock=factor(burn$Z6,labels=c("NotBurned","Burned")),
  SiteTrunk=factor(burn$Z7,labels=c("NotBurned","Burned")),
  SiteUpperLeg=factor(burn$Z8,labels=c("NotBurned","Burned")),
  SiteLowerLeg=factor(burn$Z9,labels=c("NotBurned","Burned")),
  SiteRespTract=factor(burn$Z10,labels=c("NotBurned","Burned")),
  BurnSource=factor(burn$Z11,labels=c("Chemical","Scald","Electric","Flame"))
)
burn <- tmerge(burn0, burn0, Obs, tstop=T3)
# set time to Excision
burn <- tmerge(burn, burn0, Obs, Excis=tdc(T1))
# set time to Antibiotics
burn <- tmerge(burn, burn0, Obs, Antib=tdc(T2))
# status at the end, either they had an infection or they were censored
burn$status <- as.integer(with(burn, (tstop == T3 & D3)))
# preview the data
head(burn)
```

Create a survival object with time independent covariates.
```{r}
surv1 <- Surv(burn0$T3, burn0$D3)
```

Create a survival object with time dependent covariates.
```{r}
surv2 <- Surv(time=burn$tstart, time2=burn$tstop,
              event=burn$status, type="counting")
```

Construct the palette used in the plots.
```{r}
library(scales)
library(ggsci)
mypalette = c(pal_jama("default")(7), "#003C67FF", "#7C878EFF")
show_col(mypalette)
```

## Dataset

Table 2.
```{r}
summary(burn0[burn0$Treatment=='Routine',])
sum(burn0$D1[burn0$Treatment=='Routine'])
sum(burn0$D2[burn0$Treatment=='Routine'])
summary(burn0[burn0$Treatment=='Cleansing',])
sum(burn0$D1[burn0$Treatment=='Cleansing'])
sum(burn0$D2[burn0$Treatment=='Cleansing'])
```

Wilcoxon rank-sum test.
```{r}
wilcox.test(burn0$PercentBurn[burn0$Treatment=='Cleansing'],
            burn0$PercentBurn[burn0$Treatment=='Routine'],
            alternative='less')
```

Figure 1.
```{r}
library(tidyverse)

df_group1 <- burn0 |>
  select(SiteHead, SiteButtock, SiteTrunk, SiteUpperLeg, SiteLowerLeg, SiteRespTract, Treatment) |>
  group_by(Treatment) |>
  summarise(n = n(),
            n_head = sum(SiteHead=="Burned"),
            n_but = sum(SiteButtock=="Burned"),
            n_trunk = sum(SiteTrunk=="Burned"),
            n_upleg = sum(SiteUpperLeg=="Burned"),
            n_lowleg = sum(SiteLowerLeg=="Burned"),
            n_resp = sum(SiteRespTract=="Burned"),
            .groups="drop_last") |>
  group_by(Treatment) |>
  mutate(r_head = n_head / sum(n),
         r_but = n_but / sum(n),
         r_trunk = n_trunk / sum(n),
         r_upleg = n_upleg / sum(n),
         r_lowleg = n_lowleg / sum(n),
         r_resp = n_resp / sum(n)) 

df_group2 <- burn0 |>
  select(BurnSource, Treatment) |>
  group_by(BurnSource, Treatment) |>
  summarise(n = n(), .groups="drop_last") |>
  ungroup(BurnSource) |>
  group_by(Treatment) |>
  mutate(ratio = n / sum(n)) 

df_group3 <- data.frame(Treatment=rep(c(0, 1), 6),
                        BurnSite=c(0,0,1,1,2,2,3,3,4,4,5,5),
                        Ratio=c(as.matrix(df_group1[, 9:14])))

df_group3$Treatment <- factor(df_group3$Treatment,labels=c("Routine","Cleansing"))
df_group3$BurnSite <- factor(df_group3$BurnSite,labels=c("Head","Buttock", "Trunk","UpperLeg","LowerLeg","RespTract"))

ggplot(df_group3, aes(
    x = BurnSite, 
    y = Ratio,
    fill = Treatment)) +
  geom_col(position = position_dodge2(padding = 0.25)) + 
  scale_y_continuous(expand=c(0, 0, 0, 0.1)) + 
  labs(x="Burn Site", y="Ratio") +
  theme_bw()  +
  scale_color_manual(values=mypalette) +
  scale_fill_manual(values=mypalette) +
  theme(legend.position=c(0.15, 0.8))

ggplot(df_group2, aes(
    x = BurnSource, 
    y = ratio,
    fill = Treatment)) +
  geom_col(position = position_dodge2(padding = 0.25)) + 
  scale_y_continuous(expand=c(0, 0, 0, 0.1)) + 
  labs(x="Source of Burn", y="Ratio") +
  theme_bw() +
  scale_color_manual(values=mypalette) +
  scale_fill_manual(values=mypalette) +
  theme(legend.position=c(0.15, 0.8))
```

Figure 2.
```{r}
library(survminer)
library(patchwork)
# days till Excision
surv.Excis <- Surv(burn0$T1, burn0$D1)
km.Excis <- survfit(surv.Excis~Treatment, data=burn0)
print(km.Excis)
p1 = ggsurvplot(km.Excis,
                surv.median.line="hv",
                palette="jama",
                pval=TRUE,
                pval.size=4,
                risk.table=T,
                conf.int=T,
                tables.y.text=F,
                risk.table.fontsize=3,
                tables.height=0.2,
                ggtheme=theme_bw(),
                tables.theme=theme_test() +
                  theme(axis.title=element_blank(),
                        axis.text.x=element_blank(),
                        axis.ticks.x=element_blank()),
                xlim=c(0, 52),
                ylab="Percent Not Excised",
                ylim=c(0, 1.05),
                axes.offset=F,
                legend = c(0.85, 0.8),
                legend.title = "Treatment", 
                legend.labs = c("Routine", "Cleansing"))
# days till Antib
surv.Antib <- Surv(burn0$T2, burn0$D2)
km.Antib <- survfit(surv.Antib~Treatment, data=burn0)
print(km.Antib)
p2 = ggsurvplot(km.Antib,
                surv.median.line="hv",
                palette="jama",
                pval=TRUE,
                pval.size=4,
                risk.table=T,
                tables.y.text=F,
                risk.table.fontsize=3,
                tables.height=0.2,
                conf.int=T,
                ggtheme=theme_bw(),
                tables.theme=theme_test() +
                  theme(axis.title=element_blank(),
                        axis.text.x=element_blank(),
                        axis.ticks.x=element_blank()),
                xlim=c(0, 65),
                ylab="Percent Not Treated with Antibiotics",
                ylim=c(0, 1.05),
                axes.offset=F,
                legend = c(0.85, 0.8),
                legend.title = "Treatment", 
                legend.labs = c("Routine", "Cleansing"))
```

Figure 3.
```{r}
ggsurvplot(survfit(surv.Excis~BurnSource, data=burn0),
           pval=T, 
           pval.size=4,
           palette="jama",
           ggtheme=theme_bw(),
           xlim=c(0, 52),
           ylab="Percent Not Excised",
           ylim=c(0, 1.05),
           axes.offset=F,
           legend = c(0.85, 0.7),
           legend.title = "Burn Source", 
           legend.labs = c("Chemical","Scald","Electric","Flame"))
```

## Nonparametric Estimators

Figure 4 and log-rank tests.
```{r}
p1 = ggsurvplot_combine(
  list(K_M = survfit(surv2~Treatment, data=burn),
       N_A = survfit(surv2~Treatment, type="fleming-harrington", data=burn)), 
  burn,
  palette=c("#D67C35", "#2A3D43", "#D67C36", "#2A3D44"),
  ggtheme=theme_bw(),
  censor=F,
  linetype=c(1,1,3,3),
  size=0.8,
  axes.offset=F,
  ylim=c(0, 1.05),
  ylab="Percent Infection-free",
  xlim=c(0, 102),
  legend = c(0.16, 0.28),
  legend.title = "Treatment", 
  legend.labs = c("KM:Routine", "KM:Cleansing",
                  "NA:Routine", "NA:Cleansing"))

print(survfit(surv2~Treatment, data=burn))
survdiff(surv1~Treatment, data=burn0)

p2 = ggsurvplot_combine(
  list(K_M = survfit(surv2~Race, data=burn),
       N_A = survfit(surv2~Race, type="fleming-harrington", data=burn)), 
  burn,
  palette=c("#D67C35", "#2A3D43", "#D67C36", "#2A3D44"),
  ggtheme=theme_bw(),
  censor=F,
  linetype=c(1,1,3,3),
  size=0.8,
  axes.offset=F,
  ylim=c(0, 1.05),
  ylab="Percent Infection-free",
  xlim=c(0, 102),
  legend = c(0.16, 0.28),
  legend.title = "Race", 
  legend.labs = c("KM:NonWhite", "KM:White",
                  "NA:NonWhite", "NA:White"))

print(survfit(surv2~Race, data=burn))
survdiff(surv1~Race, data=burn0)

p3 = ggsurvplot_combine(
  list(K_M = survfit(surv2~Gender, data=burn),
       N_A = survfit(surv2~Gender, type="fleming-harrington", data=burn)), 
  burn,
  palette=c("#D67C35", "#2A3D43", "#D67C36", "#2A3D44"),
  ggtheme=theme_bw(),
  censor=F,
  linetype=c(1,1,3,3),
  size=0.8,
  axes.offset=F,
  ylim=c(0, 1.05),
  ylab="Percent Infection-free",
  xlim=c(0, 102),
  legend = c(0.16, 0.28),
  legend.title = "Gender", 
  legend.labs = c("KM:Male", "KM:Female",
                  "NA:Male", "NA:Female"))

print(survfit(surv2~Gender, data=burn))
survdiff(surv1~Gender, data=burn0)

p4 = ggsurvplot(survfit(surv1~BurnSource, data=burn0), 
           palette="jama",
           ggtheme=theme_bw(),
           axes.offset=F,
           ylim=c(0, 1.05),
           ylab="Percent Infection-free",
           xlim=c(0, 102),
           legend = c(0.85, 0.75),
           legend.title = "BurnSource", 
           legend.labs = c("Chemical","Scald","Electric","Flame")
           )

print(survfit(surv2~BurnSource, data=burn))
survdiff(surv1~BurnSource, data=burn0)

p5 = ggsurvplot_combine(
  list(K_M = survfit(surv2~Excis, data=burn),
       N_A = survfit(surv2~Excis, type="fleming-harrington", data=burn)), 
  burn,
  palette=c("#D67C35", "#2A3D43", "#D67C36", "#2A3D44"),
  ggtheme=theme_bw(),
  censor=F,
  linetype=c(1,1,3,3),
  size=0.8,
  axes.offset=F,
  ylim=c(0, 1.05),
  ylab="Percent Infection-free",
  xlim=c(0, 102),
  legend = c(0.16, 0.28),
  legend.title = "Excision", 
  legend.labs = c("KM:No", "KM:Yes",
                  "NA:No", "NA:Yes"))

print(survfit(surv2~Excis, data=burn))

p6 = ggsurvplot_combine(
  list(K_M = survfit(surv2~Antib, data=burn),
       N_A = survfit(surv2~Antib, type="fleming-harrington", data=burn)), 
  burn,
  palette=c("#D67C35", "#2A3D43", "#D67C36", "#2A3D44"),
  ggtheme=theme_bw(),
  censor=F,
  linetype=c(1,1,3,3),
  size=0.8,
  axes.offset=F,
  ylim=c(0, 1.05),
  ylab="Percent Infection-free",
  xlim=c(0, 102),
  legend = c(0.16, 0.28),
  legend.title = "Antibiotics", 
  legend.labs = c("KM:No", "KM:Yes",
                  "NA:No", "NA:Yes"))

p7 = ggsurvplot(survfit(surv2~Excis+Antib, data=burn), 
           palette="jama",
           ggtheme=theme_bw(),
           axes.offset=F,
           ylim=c(0, 1.05),
           ylab="Percent Infection-free",
           xlim=c(0, 102),
           legend = c(0.2, 0.25),
           legend.title = "Excision+Antibiotics", 
           # legend.labs = c("No","Yes")
           )

p8 = ggsurvplot(survfit(surv2~Treatment+Excis, data=burn), 
           palette="jama",
           ggtheme=theme_bw(),
           axes.offset=F,
           ylim=c(0, 1.05),
           ylab="Percent Infection-free",
           xlim=c(0, 102),
           legend = c(0.25, 0.25),
           legend.title = "Treatment+Excision", 
           # legend.labs = c("No","Yes")
           )

p1$plot+p2$plot+p3$plot+p4$plot

p5$plot+p6$plot+p7$plot+p8$plot
```

## Cox PH Model

Model 1: univariate model.
```{r}
cox1 <- coxph(surv1~Treatment, data=burn0)
summary(cox1)
round(exp(confint(cox1, level=0.9)), 2)
```

Model 2: demographics.
```{r}
cox2 <- coxph(surv1~Treatment+Gender+Race, data=burn0)
summary(cox2)
round(exp(confint(cox2, level=0.9)), 2)
```

Model 3: burn characteristics.
```{r}
cox3 <- coxph(surv1~Treatment+Gender+Race+PercentBurn+SiteHead+SiteButtock+SiteTrunk+SiteUpperLeg+SiteLowerLeg+SiteRespTract+BurnSource, data=burn0)
summary(cox3)
round(exp(confint(cox3, level=0.9)), 2)
```

Model 4: time-dependent variables.
```{r}
cox4 <- coxph(surv2~Treatment+Gender+Race+PercentBurn+SiteHead+SiteButtock+SiteTrunk+SiteUpperLeg+SiteLowerLeg+SiteRespTract+BurnSource+Excis+Antib, data=burn)
summary(cox4)
round(exp(confint(cox4, level=0.9)), 2)
```

Stepwise model selection.
```{r}
cox4.5 <- step(cox4)
cox.zph(cox4.5)
```

Model 5: stratify on BurnSource.
```{r}
cox5 <- coxph(surv2~Treatment+Race+Excis+strata(BurnSource), data=burn)
summary(cox5)
round(exp(confint(cox5, level=0.9)), 2)
```

Coefficients.
```{r}
library(ggstatsplot)
library(ggfortify)
ggcoefstats(x = cox5, conf.level = 0.90) +
  theme_bw() +
  labs(y="Covariate", x="Coefficient")

# autoplot(survfit(cox5, data.frame(Treatment=levels(burn$Treatment)[1],
#                                   Race=levels(burn$Race)[2],
#                                   Excis=1)),
#          conf.int = F, censor = F)

ggsurvplot_combine(
  list(Routine = survfit(cox5,
                     data.frame(Treatment=levels(burn$Treatment)[1],
                                Race=levels(burn$Race)[2],
                                Excis=1)),
       Cleansing = survfit(cox5,
                     data.frame(Treatment=levels(burn$Treatment)[2],
                                Race=levels(burn$Race)[2],
                                Excis=1))),
  burn,
  palette=c(mypalette[1:4], "#374E56FF", "#DF8F45FF", "#00A1D6FF", "#B24746FF"),
  ggtheme=theme_bw(),
  censor=F,
  linetype=c(1,1,1,1,2,2,2,2),
  size=0.6,
  axes.offset=F,
  ylim=c(0.25, 1.05),
  ylab="Percent Infection-free",
  xlim=c(0, 101),
  legend="right")

```

PH Assumption.
```{r}
burn.zph = cox.zph(cox5)
burn.zph
```

Schoenfeld Residuals.
```{r}
source("plot_zph.R")
plot_zph(burn.zph[1],
         ggtheme=theme_bw(),
         point.col=mypalette[2],
         line.col=mypalette[1])
plot_zph(burn.zph[2],
         ggtheme=theme_bw(),
         point.col=mypalette[2],
         line.col=mypalette[1])
plot_zph(burn.zph[3],
         ggtheme=theme_bw(),
         point.col=mypalette[2],
         line.col=mypalette[1])
```

Cox-Snell Residuals.
```{r}
burn.mart <- residuals(cox5, type="martingale")
burn.cs <- burn$status - burn.mart
surv.csr <- survfit(Surv(burn.cs, burn$status)~1, type="fleming-harrington")
p1 <- ggsurvplot(surv.csr,
           data=burn,
           fun="cumhaz",
           censor=F,
           conf.int=T,
           # conf.int.fill=mypalette[1],
           ggtheme=theme_bw(),
           palette=mypalette[1],
           ylab="CumHaz of Cox-Snell Residual",
           xlim=c(0, 1.13),
           ylim=c(0, 1.3),
           axes.offset=F,
           legend="none")
p.csr <- p1$plot + geom_abline(slope=1,
                               intercept = 0,
                               color=mypalette[2],
                               linewidth=1.2,
                               linetype="dashed")
```

Martingale residuals vs. the linear predictor; Deviance residuals vs. the linear predictor.
```{r}
p.mart <- ggcoxdiagnostics(cox5,
                           facet=F,
                           type="martingale",
                           point.col=mypalette[2],
                           sline.lty=1, 
                           sline.col=mypalette[1],
                           hline=F,
                           point.size=1.2,
                           sline.se.col=mypalette[2],
                           ggtheme=theme_bw(),
                           ylab="Martingale Residual",
                           xlab="Linear Predictor") + facet_null()
p.dev <- ggcoxdiagnostics(cox5,
                          type="deviance",
                          point.col=mypalette[2],
                          sline.lty=1, 
                          sline.col=mypalette[1],
                          hline=F,
                          point.size=1.2,
                          sline.se.col=mypalette[2],
                          ggtheme=theme_bw(),
                          ylab="Deviance Residual",
                          xlab="Linear Predictor") + facet_null()
```

Df Betas.
```{r}
p.dfbeta <- ggcoxdiagnostics(cox5,
                             type="dfbeta",
                             point.col=mypalette[2],
                             sline.lty=1, 
                             sline.col=mypalette[1],
                             hline=F,
                             point.size=1.2,
                             sline.se.col=mypalette[2],
                             ggtheme=theme_bw(),
                             ylab="Dfbetas",
                             xlab="Observation Order") +
  theme(strip.background = element_blank())
```

Join the plots.
```{r}
(p.csr | p.mart | p.dev) / p.dfbeta
```

Outlier detection.
```{r}
burn.mart <- residuals(cox5, type="martingale")
burn.mart[abs(burn.mart) > 1] # 51

burn.dev <- residuals(cox5, type="deviance")
burn.dev[abs(burn.dev) > 2] # 71, 108, 134, 211, 214, 231, 286

burn.dfb <- residuals(cox5, type="dfbeta")
burn.dfb[abs(burn.dfb[, 1]) > 0.05, 1] # 51, 134, 156, 211, 214, 231, 240, 286
burn.dfb[abs(burn.dfb[, 2]) > 1, 2] # 214
burn.dfb[abs(burn.dfb[, 3]) > 0.1, 3] # 129, 158, 160, 181, 240

# ROW NUMBER: 51, 134, 211, 214, 231, 286
# ID: 32, 79, 115, 116, 125 and 153
burn0[c(32, 79, 115, 116, 125, 153), -1]
```

## Aalen Additive Hazard Model

```{r}
library(timereg)
source("geom_stepribbon.R")
# vars
# Treatment+Gender+Race+BurnSource+Excis
coef_aalen <- function(aa, level=0.1){
  list(time=aa[["cum"]][, 1],
       yy=aa[["cum"]][, -1],
       ylow=aa[["cum"]][,-1]-qnorm(1-level/2)*sqrt(aa[["var.cum"]][,-1]),
       yup=aa[["cum"]][,-1]+qnorm(1-level/2)*sqrt(aa[["var.cum"]][,-1]))
}
```

Model 6.
```{r}
aa.whole1 <- aalen(surv2~Treatment+Gender+Race+BurnSource+Excis,
                  data=burn, n.sim=300, start.time=0)
summary(aa.whole1)
```

Model 7.
```{r}
library(gridExtra)
aa.whole <- aalen(surv2~const(Treatment)+Gender+Race+const(BurnSource)+Excis,
                  data=burn, n.sim=300, start.time=0)
summary(aa.whole)
coefs <- coef_aalen(aa.whole)
plots <- lapply(1:ncol(coefs$yy), function(i) {ggplot() + geom_step(mapping=aes(x=coefs$time,
                                 y=coefs$yy[, i]),
                     color=mypalette[i],
                     size=0.8) +
  geom_stepribbon(mapping=aes(x=coefs$time,
                              ymin=coefs$ylow[, i],
                              ymax=coefs$yup[, i]),
                  alpha=0.32,
                  fill=mypalette[i]) +
  theme_bw() + theme(plot.subtitle = element_text(hjust = 0.5)) +
  labs(x=element_blank(), y=element_blank(),
       subtitle=colnames(coefs$yy)[i])}) %>% do.call(grid.arrange, .)
```

Model 8.
```{r}
# cut
burn.split <- survSplit(surv2 ~ ., data=burn, cut=c(14), episode="tgroup")
# early
burn.early = burn.split[burn.split$tgroup==1, ]
aa.early <- aalen(burn.early$surv2~const(Treatment)+const(Gender)+const(Race)+const(Excis)+const(BurnSource),
                  n.sim=300, data=burn.early, start.time=0)
summary(aa.early)
coefs <- coef_aalen(aa.early)
# late
burn.late = burn.split[burn.split$tgroup==2, ]
aa.late <- aalen(burn.late$surv2~Treatment+Gender+const(Race)+const(Excis)+const(BurnSource), 
                 n.sim=300, data=burn.late, start.time=0)
summary(aa.late)
coefs <- coef_aalen(aa.late)
plots <- lapply(1:ncol(coefs$yy), function(i) {ggplot() + geom_step(mapping=aes(x=coefs$time,
                                 y=coefs$yy[, i]),
                     color=mypalette[i],
                     size=0.8) +
  geom_vline(xintercept=14, size=0.8,
             linetype="dashed", color="darkgrey") +
  geom_stepribbon(mapping=aes(x=coefs$time,
                              ymin=coefs$ylow[, i],
                              ymax=coefs$yup[, i]),
                  alpha=0.32,
                  fill=mypalette[i]) +
  theme_bw() + theme(plot.subtitle = element_text(hjust = 0.5)) +
  labs(x=element_blank(), y=element_blank(),
       subtitle=colnames(coefs$yy)[i])}) %>% do.call(grid.arrange, .)
```

